Personalised Map Matched Imputation: imputing missing data from smartphone location logs

### 2018-05-05 Thesis Manuscript/apaManu.Rmd
finishing touches

abstract: |
  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underutilised in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurments and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.

>>>>

abstract: |
  Personal mobility, or how people move in their environment, is associated with a vast range of behavioural traits and outcomes, such as socioeconomic status, personality and mental health. The widespread adoption of location-sensor equipped smartphones has generated a wealth of objective personal mobility data. Nonetheless, smartphone collected personal mobility data has remained underused in behavioural research, partly due to the practical difficulties associated with obtaining the data and partly because of the methodological complexity associated with analysing it. Recent changes in European regulation have made it easier for researchers to obtain this data, but the methodological difficulties remain. The difficulty lies in that smartphone location data is irregularly sampled, sparse and often inaccurate.  This results in a high proportion of missing data and significant noise. In this paper we present a method called Personal Map Matched Imputation (PPMI) to deal with missing data and noise in smartphone location logs. The main innovation of PPMI is that it creates a personalised spatial map for each individual based on all the available data. In doing so PPMI leverages the regularity of human mobility in order to smoothen noisy measurements and impute missing data values. By simulating missing periods in real data we find that a simple implementation of PPMI performs as well as existing methods for short (5 minute) missing intervals and substantially better for longer (1 day) missing intervals. When imputing a subset of real  missing data where travel logs are available as a reference points, we find that PPMI performs substantially better than existing models.
